ML Estimation of Matrix--$t$ Distributions with Expectation Maximization

### 2017-01-12 paper.tex
Working on section 2.

All matrices referred in this note are real. For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\left\{S_p\succ 0\right\}$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$.

We first review some basic definitions on matrix--$t$ distributions. A matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,n)$ of dimensions $p\times m$ is specified by 4 parameters: the location matrix $M$ of dimensions $p\times m$ , the row and column SPD scale matrices $\Sigma$ and $\Omega$ of dimensions $p\times p$ and $m\times m$, respectively, and the degrees of freedom parameter $n>0$. A $p \times m$ random matrix $T$ is said to be matrix--$t$ distributed $T_{p,m}(M,\Sigma,\Omega,n)$ with location parameter $M$ if given the row scale parameter $S$, $T$ is matrix normal distributed

The derivation of the EM formulae is mainly based on the integral identity \citep{gupta1999matrix}
\begin{equation}
\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}\,dS=\Gamma_{p}\left(x+\frac{1+p}{2}\right)|A|^{x+\frac{1+p}{2}}.\label{eq:intmultgammadef}
\end{equation}
Here the domain of integration is the set of real symmetric
positive definite matrices with dimension $p$, $A$ is a $p\times p$
nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma
function, given in terms of the standard gamma function $\Gamma(\cdot)$
by
\begin{equation}
\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+\frac{1-j}{2}\right).\label{eq:multgammadef}
\end{equation}
We also define the multivariate digamma function
%
\begin{equation} \psi_{p}(a)=\frac{\partial}{\partial
	a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
\end{equation}

A random $p\times m$ matrix $T$ following a matrix--$t$ distribution
$T_{p,m}(M,\Sigma,\Omega,n)$ has the probability density
\begin{equation}
f(T;\left\{M,\Sigma,\Omega,n\right\})=\frac{\Gamma_{p}\left(\frac{p+n+m-1}{2}\right)}{\Gamma_{p}\left(\frac{n+p-1}{2}\right)\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}
|I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|^{\frac{1-m-p-n}{2}},
\label{eq:matrixtpdf}
\end{equation}
where $\Sigma$ and $\Omega$ are both real symmetric positive definite
matrices of dimension $p$ and $m$, respectively, $M$ is a $p\times m$
matrix that indicates the location of the distribution, and $n$ is the
number of degrees of freedom.

Following \citet{gupta1999matrix}, we begin by deriving the
probability density of the matrix--$t$ distribution starting from
a matrix normal distributed random variate $X\sim N_{p,m}(M,S,\Omega)$
with Wishart distributed row scale parameter $S \sim W_{p}\left(\Sigma,\nu\right)$ of appropriate degrees of freedom $\nu$ (independent from $X$), and marginalizing over $S$. Note that the row scale
parameter $S$ is a real symmetric positive definite matrix. We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulae.

The $p\times m$ random matrix $X \sim N_{p,m}(M,S,\Omega)$ has density
\begin{equation}
	\frac{1}{(2\pi)^{pm/2}|\Omega|^{p/2}|S|^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right),\label{eq:ndistpdf}
\end{equation}
where $S$ and $\Omega$ are symmetric positive definite matrices of dimension $p$ and $m$, respectively, and $M$ (the expectation of $X$) is a $p\times m$ matrix. The matrix $S$ is proportional to the covariance of the rows of $X$ while the matrix $\Omega$ is proportional to the covariance of the columns of $X$. In turn the $p\times p$ matrix $S\sim W_{p}\left(\Sigma,\nu\right)$ has density
\begin{equation}
\frac{|S|^{\frac{\nu-p-1}{2}}}{2^{\nu p/2}|\Sigma|^{\nu/2}\Gamma_{p}(\frac{\nu}{2})}\exp\left(-\textrm{tr}\left[\Sigma^{-1}S\right]\right),\label{eq:wdistpdf}
\end{equation}
where $\Sigma$ is a $p\times p$ symmetric positive definite matrix, $\nu>p-1$
is the number of degrees of freedom, and $\Gamma_{p}$ is the
multivariate gamma function defined in Eq.~\eqref{eq:multgammadef}.

>>>>

All matrices referred in this note are real. For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\left\{S_p\succ 0\right\}$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$. The multivariate gamma
function, $\Gamma_{p}(\cdot)$, given in terms of the standard gamma function $\Gamma(\cdot)$ has the form
\begin{equation}
\Gamma_{p}(x)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(x+\frac{1-j}{2}\right).\label{eq:multgammadef}
\end{equation}
We also define the multivariate digamma function
%
\begin{equation}
\psi_{p}(x):=\frac{\partial}{\partial x}\log\Gamma_{p}(x)=\sum_{j=1}^{p}\psi\left(x+\frac{1-j}{2}\right).
\end{equation}

We first review some basic definitions on matrix--$t$ distributions. A matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ of dimensions $p\times m$ is specified by 4 parameters: the location matrix $M$ of dimensions $p\times m$ , the row and column SPD scale matrices $\Sigma$ and $\Omega$ of dimensions $p\times p$ and $m\times m$, respectively, and the degrees of freedom parameter $\nu>0$. A $p \times m$ random matrix $T$ is said to be matrix--$t$ distributed $T_{p,m}(M,\Sigma,\Omega,\nu)$ with location parameter $M$ if given the scale parameter $S$, a SPD matrix of dimensions $p\times p$, the random matrix $T$ is matrix normal distributed $N_{p,m}(M,S,\Omega)$ and $S$ is itself inverse Wishart distributed $W^{-1}(\Sigma,\nu +p -1)$:
\begin{equation}
T|M,\Sigma,\Omega,\nu,S \sim N_{p,m}(M,S,\Omega)\label{eq:tdef1}
\end{equation}
and
\begin{equation}
S|\Sigma,\nu \sim W^{-1}(\Sigma,\nu+p-1).\label{eq:tdef2}
\end{equation}

The matrix normal distribution $N_{p,m}(M,S,\Omega)$ has density
\begin{equation}
\frac{1}{(2\pi)^{pm/2}|\Omega|^{p/2}|S|^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
\end{equation}
at $X$ while the inverse Wishart distribution $W^{-1}(\Sigma,n)$ with $n>p-1$ has density
\begin{equation}
\frac{|\Sigma|^{\frac{n}{2}}}{2^{\frac{n p}{2}}|S|^{\frac{n+p+1}{2}}\Gamma_{p}(\frac{n}{2})}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Sigma S^{-1}\right]\right)\label{eq:wdistpdf}
\end{equation}
at $S$. Here $\Gamma_{p}(\cdot)$ is the
multivariate gamma function defined in Eq.~\eqref{eq:multgammadef}.

Follow the definition in Eqs.~\eqref{eq:tdef1} and \eqref{eq:tdef2} and the densities given above we can see the density function $f(T;\left\{M,\Sigma,\Omega,\nu\right\})$ of a matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ satisfies
\begin{equation}
\frac{\Gamma_{p}\left(\frac{p+\nu+m-1}{2}\right)}{\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}
|I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|^{\frac{1-m-p-\nu}{2}}
\label{eq:matrixtpdf}
\end{equation}
with $I_p$ the identity matrix of dimension $p$. The marginalization over $S$ can be obtained by using the Jacobain determinant
\begin{equation}
...
\end{equation}
and the integral identity \citep{gupta1999matrix}
\begin{equation}
\int_{\left\{S_p\succ 0\right\}}e^{-\textrm{tr}[S_pA]}|S_p|^{\alpha}\,dS_p=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{\alpha+\frac{1+p}{2}},\label{eq:intmultgammadef}
\end{equation}
where $A$ is a $p\times p$ nonsingular matrix.

We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulae.
